\ismasec{Summary and Conclusions}
Overall, basic DSP was reviewed with a focus on 
\emph{translation-invariance} -- translation-invariant
operators of $\LG$, and translation-invariant
subspaces of $\LG$. 
When such great significance is attached to translation-invariance, 
a deeper understanding of exponential functions, and their 
unrivaled status in classical DSP, is possible.  
In particular, exponentials are the \emph{characters} 
of abelian group indexing sets, such as $\Z/N$, over which
classical DSP is performed.  
Each character of an abelian group represents a one-dimensional
translation-invariant subspace, and the characters are eigenvectors
of translations, therefore, of convolutions. 
%(which explains why expansions in the character basis -- \eg
%classical Fourier transform -- diagonalize the convolution
%product).  

We described the group algebra $\CG$,
the algebra isomorphism $\LG \simeq \CG$, and why it is
useful for manipulations involving (generalized)
translations and convolutions of the space of signals. 
We saw that, for an abelian group $A$, translations of $\LA$
represent simple linear shifts in space or time, while for a nonabelian
group $G$, translations of $\LG$ are more general than
simple spatial or temporal shifts.  This leads to
more interesting translation-invariant subspaces.

Motivating many studies in the area of noncommutative
harmonic analysis (including this one) is a simple
but important fact about the generalized translations that
result when a signal is indexed by a nonabelian group.
As we have seen, such operations offer more complex and
interesting signal transformations.  Equally important,
however, is the fact that each transformation can be written
as a left-multiplication. Thus, the increase in signal transform
complexity resulting from a nonabelian indexing scheme
comes at no increase in computational complexity. 


