\frame[label=milestones]{
    \frametitle{Open Problems}

\begin{enumerate}[1.]
\item
\label{item:1} Are homomorphic images of a representable lattices representable?

\bigskip

\item
\label{item:2} Are subdirect products of representable lattices representable?

\bigskip

\item\label{item:3} Does representable imply ``group representable?''  \\[4pt]
{\small i.e., is the congruence lattice of a finite algebra isomorphic to an interval in the subgroup lattice of a finite group?}

\bigskip
%% \item For which $n$ is $M_n$ a congruence lattice of a finite algebra?
%% \item\label{item:4} Suppose $L$ is group representable.\\[4pt]
%% Is the filter+ideal lattice $\tilde{L} = \alpha^\uparrow \cup \beta^\downarrow$ group representable, for all $\alpha, \beta \in L$.

%% \bigskip

\item\label{item:5} Is the class of representable lattices recursive?

\end{enumerate}
}

%% \frame[label=milestones]{
%%     \frametitle{Remarks on the problems}
%% \begin{theorem}[Lucchini, 1994] The lattice $M_n$ occurs as an interval in a subgroup lattices of a finite group when
%% \[
%% n = q + 2 \qquad \text{ or } \qquad n =
%% \frac{q^t + 1}{q + 1}+ 1,\]
%% where $q$ is a prime power and $t$ is an odd prime.
%% \end{theorem}
%% At present the smallest cases for which no occurrence of $M_n$ is known are $n = 16,
%% 23, 35, \dots$.
%% }

\frame[label=milestones]{
    \frametitle{Remarks on the problems}
    \begin{itemize}
    \item 
Are homomorphic images of representable lattices representable?
\\[4pt]
If $L = \Con \<A, F\>$ and $\tilde{L}$ is the homomorphic image $\{\theta \cap B^2 \mid \theta \in L\}$, where $B = e(A)$ for some operation $e^2 = e \in F$, then $\tilde{L}=\Con \<B, F\!\mid_B\>$ .

\bigskip
\uncover<2->{
\item 
Is the class of representable lattices recursive?
\\[4pt]
In other words, is the membership problem for this class decidable?
`\\[4pt]
(Note that the class is recursively enumerable by the closure method.)
\\[4pt]
\uncover<3->{
This question asks whether it's possible to write a program that, when given a finite lattice $L$, halts with output True if $L$ is representable and False otherwise.
\\[4pt]

A negative answer would solve the finite lattice representation problem.}
}
    \end{itemize}
}

\begin{frame}[fragile]
    \frametitle{Open Problems}
One approach: try to find a computable function $f$ such that, if $L$ is a representable lattice of size $n$, then $L$ is representable as the congruence lattice of an algebra of cardinality $f(n)$ or smaller.
\\[8pt]
%% which takes as input a representable finite lattice $L$, and outputs a number $f(L)$ such that $L$ is representable as a congruence lattice on an algebra of size at most $f(L)$.

This would answer the decidability question, as follows:
\begin{small}
\begin{verbatim}
IsRepresentable( L ) {

    n:= Size( L )
    N:= f( n )

    for each L' in Sub[Eq(N)] {

        if ( L' isomorphic to L ) and ( L' is closed ):
            return True

    }

    return False

}
\end{verbatim}
\end{small}
\end{frame}

